\chapter*{Contribution to publications}
\addcontentsline{toc}{chapter}{Contribution to publications}
\fancyhead[L]{{\color{gray}\nouppercase{Contribution to publications}}}

\setlength{\parindent}{0pt}
\setlength{\parskip}{3pt}

This thesis is presented as a manuscript-based thesis. Each manuscript contains important contributions from my co-authors, without whom this work would not have been possible. Below, I describe individual contributions to each manuscript that together constitute the main chapters of my thesis.

%%
\vspace{24pt}
%%

\textsc{\secbfcolor{Chapter 2}} -- \textsc{Quantifying immunoregulation by autoantigen-specific T-regulatory type 1 cells in mice with simultaneous hepatic and extra-hepatic autoimmune disorders}

\hspace*{0.25in}%
\begin{minipage}{6in}
    \setlength{\parskip}{1pt}
    \vspace{6pt}
    Hassan Jamaleddine$^{1 }$, Pere Santamaria$^{2,3}$, Anmar Khadra$^{1}$

    \vspace{6pt}
    
    {\small $^1$~Department of Physiology, McGill University, Montreal, Canada \par $^2$~Department of Microbiology, Immunology \& Infectious Diseases, Calgary, Canada \par $^3$~Institut D’Investigacions Biom\`{e}diques August Pi i Sunyer, Carrer del Rossell\'{o}, Barcelona, Spain}

    \vspace{6pt}
    
    \textit{Conceptualization}: PS and AK. \textit{Methodology}: HJ. Software: HJ. \textit{Formal analysis}: HJ. \textit{Investigation}: HJ. \textit{Resources}: AK. \textit{Data curation}: HJ. \textit{Writing -- original draft}: HJ. \textit{Writing – review and editing}: HJ, PS, and AK. \textit{Visualization}: HJ. \textit{Supervision}: AK. \textit{Project administration}: AK. \textit{Funding acquisition}: PS and AK.

    \vspace{6pt}
    
    Status: published in \textit{Immunology}, 2020.
\end{minipage}

%%
\vspace{24pt}
%%


\textsc{\secbfcolor{Chapter 3}} -- \textsc{Chronic infection control relies on T cells with lower foreign antigen binding strength generated by N-nucleotide diversity}

\hspace*{0.25in}%
\begin{minipage}{6in}
    \setlength{\parskip}{1pt}
    \vspace{6pt}
    Hassan Jamaleddine$^{1 \dagger}$, Dakota Rogers$^{1,2 \dagger }$, Geneviève Perreault$^{2}$, Dhanesh Patel$^{1,2}$, Judith N. Mandl$^{1,2,3 \ddagger }$, Anmar Khadra$^{1 \ddagger }$

    \vspace{6pt}
    
    {\small $^1$~Department of Physiology, McGill University, Montreal, Canada \par $^2$~McGill University Research Centre on Complex Traits, McGill University, Montreal, Canada \par $^3$~Department of Microbiology and Immunology, McGill University, Montreal, Canada \par $^{\dagger,\,\ddagger}$ Equal contribution}
\end{minipage}

\hspace*{0.25in}%
\begin{minipage}{6in}
    \textit{Conceptualization}: JNM and AK. \textit{Methodology}: HJ (modelling) and DR (experiments). \textit{Software}: HJ. \textit{Formal analysis}: HJ and DR. \textit{Investigation}: HJ, DR, GP, and DP. \textit{Resources}: JNM and AK. \textit{Data curation}: HJ and DR. \textit{Writing -- original draft}: HJ and DR. \textit{Writing – review and editing}: HJ, DR, GP, DP, JNM, and AK. \textit{Visualization}: HJ, DR, JNM, and AK. \textit{Supervision}: JNM and AK. \textit{Project administration}: JNM and AK. \textit{Funding acquisition}: JNM and AK.
    
    \vspace{6pt}

    Status: submitted, under review (final version subject to change).
\end{minipage}

%%
\vspace{24pt}
%%


\textsc{\secbfcolor{Chapter 4}} -- \textsc{T cell receptor repertoires with reduced N-nucleotide diversity may be more susceptible to pathogen immune escape}

\hspace*{0.25in}%
\begin{minipage}{6in}
    \setlength{\parskip}{1pt}
    \vspace{6pt}
    Hassan Jamaleddine$^{1}$, Judith N. Mandl$^{1,2,3 \dagger }$, Anmar Khadra$^{1 \dagger }$

    \vspace{6pt}
    
    {\small $^1$~Department of Physiology, McGill University, Montreal, Canada \par $^2$~McGill University Research Centre on Complex Traits, McGill University, Montreal, Canada \par $^3$~Department of Microbiology and Immunology, McGill University, Montreal, Canada \par $^{\dagger}$ Equal contribution}

    \vspace{6pt}
    
    \textit{Conceptualization}: HJ, JNM, and AK. \textit{Methodology}: HJ. \textit{Software}: HJ. \textit{Formal analysis}: HJ. \textit{Investigation}: HJ. \textit{Resources}: JNM and AK. \textit{Data curation}: HJ. \textit{Writing -- original draft}: HJ. \textit{Writing – review and editing}: HJ, JNM, and AK. \textit{Visualization}: HJ. \textit{Supervision}: JNM and AK. \textit{Project administration}: JNM and AK. \textit{Funding acquisition}: JNM and AK.

    \vspace{6pt}

    Status: in preparation (final version subject to change).
\end{minipage}

\setlength{\parindent}{16pt}
\setlength{\parskip}{6pt}